---
title: "GSEA waterfall, volcano, and comparative NES plots"
description: "Standalone R functions for plotting precomputed GSEA results."
site: bookdown::bookdown_site
documentclass: book
---

```{r tutorial-options, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
```

<div class="tutorial-title-block">
<h1 class="title">GSEA waterfall, volcano, and comparative NES plots</h1>
<div class="tutorial-subtitle">Standalone R functions for visualizing ranked GO biological-process enrichment results.</div>

<div class="tutorial-meta-block">
<div><span>Author:</span> Zoheb Khan</div>
<div><span>Modified:</span> 2026-08-01</div>
<div><span>Compiled:</span> 2026-08-01</div>
<div><span>Environment:</span> R version 4.6.1 (2026-06-24)</div>
<div><span>Bioconductor:</span> 3.23</div>
<div><span>Example:</span> GSE122380 GO biological-process GSEA contrasts</div>
<div><span>Source:</span> <a href="https://github.com/ZohebKhan1/gsea-waterfall">GitHub repository</a></div>
</div>
</div>

# Overview

`gsea-waterfall` visualizes **precomputed** Gene Set Enrichment Analysis (GSEA) results as ranked NES waterfalls, symmetric or directional volcano plots, and cross-contrast NES scatterplots. The functions return ggplot objects without running GSEA, changing supplied statistics, or writing files.

The examples use fixed GO biological-process results from GSE122380 cardiomyocyte differentiation contrasts. Plot controls select, group, and label displayed rows; significance classes use the explicit adjusted-p-value cutoff without altering the supplied statistics.

The code below is the user-facing workflow. The exact export settings used to
produce the bundled SVG files are maintained in `R/01_generate_gsea_figures.R`;
the tutorial keeps those export-only styling details out of the main workflow.
The figures below are the bundled examples produced by that export script, so
their labels and category seeds may be more specific than the compact calls
shown here.

The waterfall layout is inspired by pathway-level figures in [Ciceri, *Nature*, 2024](https://doi.org/10.1038/s41586-023-06984-8), [Xu, *Nature Cell Biology*, 2025](https://doi.org/10.1038/s41556-025-01751-5), [Vuong, *Science*, 2026](https://doi.org/10.1126/science.aea1259), and [Risgaard, *Science*, 2026](https://doi.org/10.1126/science.aea1549).

# Install and source

Install the plotting dependencies once:

```r
install.packages(c('ggplot2', 'ggrepel', 'RColorBrewer', 'scales'))
```

Download the shared utilities first, then the plot modules:

```r
base_url <- 'https://raw.githubusercontent.com/ZohebKhan1/gsea-waterfall/main/functions'

download.file(paste0(base_url, '/gsea_plot_utils.R'), 'gsea_plot_utils.R')
download.file(paste0(base_url, '/gsea_waterfall_plot.R'), 'gsea_waterfall_plot.R')
download.file(paste0(base_url, '/gsea_volcano_plot.R'), 'gsea_volcano_plot.R')
download.file(paste0(base_url, '/gsea_half_volcano_plot.R'), 'gsea_half_volcano_plot.R')
download.file(paste0(base_url, '/gsea_nes_scatter_plot.R'), 'gsea_nes_scatter_plot.R')

source('gsea_plot_utils.R')
source('gsea_waterfall_plot.R')
source('gsea_volcano_plot.R')
source('gsea_half_volcano_plot.R')
source('gsea_nes_scatter_plot.R')
```

# Input contract

Each table represents one GSEA contrast with one row per unique term. The default columns are:

| Column | Required | Meaning |
|:---|:---|:---|
| `go_description` | yes | GO term description and fallback matching key |
| `go_term_id` | no | stable matching key used automatically when present |
| `NES` | yes | normalized enrichment score |
| `pval` | yes | nominal GSEA p-value in `[0, 1]` |
| `padj` | yes | adjusted p-value in `[0, 1]` |

Required columns must be complete, and NES values must be finite. Key precedence is explicit `id_col`, then `go_term_id` when present, otherwise `go_description`. Comparative scatterplots use the exact shared-key intersection and do not impute absent terms.

If your result table uses different column names, map them when reading it:

```r
fgsea_results <- read_gsea_result_csv(
  'fgsea_results.csv',
  term_col = 'pathway',
  nes_col = 'NES',
  pvalue_col = 'pval',
  padj_col = 'padj'
)
```

For comparative plots, use `id_col` with a stable term identifier whenever one
is available. Without stable IDs, term descriptions must match exactly across
tables.

Load the bundled examples from the repository root:

```r
gsea_cardiomyocyte_vs_mesoderm <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_cardiomyocyte_vs_mesoderm.csv'
)
gsea_day9_vs_day6 <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_day9_vs_day6.csv'
)
gsea_day3_vs_day1 <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_day3_vs_day1.csv'
)

print(head(gsea_cardiomyocyte_vs_mesoderm, 5L))
```

# Ranked waterfall plots

`plot_gsea_waterfall()` selects one NES direction, ranks by `NES`, `padj`, `pvalue`, or `pval`, and retains `top_n` terms. `term_groups` controls colors and `label_terms` controls labels. Matches are case-insensitive against term descriptions or IDs.

## Positive NES waterfall

```r
positive_term_groups <- list(
  'Cardiac development' = c('cardiac', 'heart', 'muscle'),
  'Metabolism' = c('mitochondrial', 'metabolic', 'oxidative')
)

positive_waterfall_plot <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'positive',
  top_n = 100L,
  rank_by = 'NES',
  label_n = 12L,
  term_groups = positive_term_groups
)
```

`direction` chooses the NES side to display. `top_n` is applied after that
filter and after ranking. Terms are neutral unless `term_groups` is supplied;
each seed is matched case-insensitively against term IDs or descriptions.

```{r positive-waterfall-figure, results='asis', out.width='80%', fig.alt='Positive NES waterfall for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-positive-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_positive.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 1.</strong> Positive NES terms ranked by NES, with selected term groups and labels.</div>

## Negative NES waterfall

```r
negative_waterfall_plot <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'negative',
  top_n = 100L,
  rank_by = 'NES',
  label_n = 10L
)
```

Use `rank_by = 'padj'` or `rank_by = 'pvalue'` when statistical significance,
rather than NES magnitude, should determine the displayed order. `label_n`
selects labels automatically; use `label_terms` when specific terms should be
shown.

```{r negative-waterfall-figure, results='asis', out.width='80%', fig.alt='Negative NES waterfall for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-negative-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_negative.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 2.</strong> Bundled negative NES example with caller-defined ribosomal, mitotic, and DNA-replication term groups.</div>

# Volcano plots

Use `plot_gsea_volcano()` for both NES directions in one panel and `plot_gsea_half_volcano()` for a directional view. All examples use the established adjusted p-value cutoff of 0.05.

## Symmetric volcano

```r
symmetric_volcano_plot <- plot_gsea_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  padj_cutoff = 0.05,
  label_n = 10L,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

`padj_cutoff` controls significance coloring and the displayed counts. It does
not remove rows from the plot. Use `label_terms` instead of `label_n` when you
want to label exact terms.

```{r symmetric-volcano-figure, results='asis', out.width='58%', fig.alt='Symmetric GSEA volcano for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-symmetric-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_volcano_cardiomyocyte_vs_mesoderm.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 3.</strong> Symmetric NES volcano colored by adjusted-p-value significance and direction.</div>

## Directional half-volcano plots

```r
positive_half_volcano_plot <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'positive',
  p_col = 'padj',
  padj_cutoff = 0.05,
  label_n = 10L,
  term_groups = positive_term_groups,
  inner_nes_limit = 1,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

`inner_nes_limit = 1` keeps terms with positive NES at least 1 in magnitude.
Set `p_col = 'pvalue'` to show nominal p-values instead; significance coloring
continues to use `padj_cutoff`.

```{r positive-half-volcano-figure, results='asis', out.width='72%', fig.alt='Positive NES half-volcano for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-positive-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_positive.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 4.</strong> Bundled positive NES half-volcano with caller-defined term groups and labels.</div>

```r
negative_half_volcano_plot <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'negative',
  p_col = 'padj',
  padj_cutoff = 0.05,
  label_n = 10L,
  inner_nes_limit = 1,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

```{r negative-half-volcano-figure, results='asis', out.width='72%', fig.alt='Negative NES half-volcano for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-negative-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_negative.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 5.</strong> Bundled negative NES half-volcano with caller-defined term groups and labels.</div>

# Comparative NES scatterplot

`plot_gsea_nes_scatter()` compares exact shared term keys. By default it keeps
terms significant in at least one contrast and does not draw a fitted line. The
line is descriptive when requested and does not alter enrichment statistics.

```r
nes_scatter_plot <- plot_gsea_nes_scatter(
  gsea_x = gsea_day9_vs_day6,
  gsea_y = gsea_day3_vs_day1,
  x_label = 'Day 9 vs. Day 6 NES',
  y_label = 'Day 3 vs. Day 1 NES',
  color_by = 'significance',
  quadrant = 'all',
  x_name = 'Day 9 vs. Day 6',
  y_name = 'Day 3 vs. Day 1',
  label_n_x_only = 3L,
  label_n_y_only = 3L,
  label_n_both = 3L,
  padj_cutoff = 0.05,
  include_nonsignificant = TRUE,
  equal_axis_limits = TRUE,
  show_fit_line = TRUE
)
```

```{r all-quadrants-scatter-figure, results='asis', out.width='100%', fig.alt='Comparative NES scatterplot for Day 9 versus Day 6 and Day 3 versus Day 1'}
cat('<div id="fig-all-quadrants-scatter" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_scatter_day9_vs_day6_x_day3_vs_day1_all_quadrants.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 6.</strong> Shared terms colored by significance in one, both, or neither contrast.</div>

Set `include_nonsignificant = FALSE` to focus on terms significant in at least
one contrast. Use `quadrant = 'q1'` or `quadrant = 'q3'` to focus on concordant
positive or negative NES values. `equal_axis_limits = TRUE` makes the two NES
axes directly comparable.

# API summary

| Function | Contract |
|:---|:---|
| `read_gsea_result_csv()` | reads and validates one precomputed CSV without filtering rows |
| `plot_gsea_waterfall()` | returns a ranked one-direction ggplot |
| `plot_gsea_volcano()` | returns a symmetric NES-versus-adjusted-p-value ggplot |
| `plot_gsea_half_volcano()` | returns a directional ggplot after the explicit NES boundary |
| `plot_gsea_nes_scatter()` | returns a ggplot from the exact shared term-key intersection |

Source documentation defines the complete argument contracts, including the common `term_col`, `nes_col`, `pvalue_col`, `padj_col`, and optional `id_col` arguments.

# References

<ol class="reference-list">
<li>Subramanian A, Tamayo P, Mootha VK, et al. Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. <em>PNAS</em>. 2005. <a href="https://www.pnas.org/doi/10.1073/pnas.0506580102">https://www.pnas.org/doi/10.1073/pnas.0506580102</a></li>
<li>Ciceri G, Baggiolini A, Cho HS, et al. An epigenetic barrier sets the timing of human neuronal maturation. <em>Nature</em>. 2024;626:881-890. <a href="https://doi.org/10.1038/s41586-023-06984-8">https://doi.org/10.1038/s41586-023-06984-8</a></li>
<li>Xu N, Cho HS, Hackland JOS, et al. Genome-wide CRISPR screen identifies Menin and SUZ12 as regulators of human developmental timing. <em>Nature Cell Biology</em>. 2025;27:1411-1421. <a href="https://doi.org/10.1038/s41556-025-01751-5">https://doi.org/10.1038/s41556-025-01751-5</a></li>
<li>Vuong CK, Weber A, Seong P, et al. A single-cell multiomic analysis identifies molecular and gene-regulatory mechanisms dysregulated in developing Down syndrome neocortex. <em>Science</em>. 2026;392:eaea1259. <a href="https://doi.org/10.1126/science.aea1259">https://doi.org/10.1126/science.aea1259</a></li>
<li>Risgaard RD, et al. Molecular and cellular processes disrupted in the early postnatal Down syndrome prefrontal cortex. <em>Science</em>. 2026;392:eaea1549. <a href="https://doi.org/10.1126/science.aea1549">https://doi.org/10.1126/science.aea1549</a></li>
</ol>
