---
title: "Comparative ranked GSEA waterfall, volcano, scatter-style plots"
description: "R functions for plotting ranked GSEA GO terms and comparative cross-contrast NES scatterplots."
site: bookdown::bookdown_site
documentclass: book
---

```{r tutorial-meta, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)

tutorial_dependency_packages <- c(
  "base64enc",
  "BiocManager",
  "bookdown",
  "ggplot2",
  "ggrepel",
  "GO.db",
  "jsonlite",
  "RColorBrewer",
  "scales",
  "svglite",
  "systemfonts"
)
tutorial_dependency_versions <- vapply(
  tutorial_dependency_packages,
  function(pkg) {
    if (!requireNamespace(pkg, quietly = TRUE)) {
      return("not installed")
    }
    as.character(utils::packageVersion(pkg))
  },
  character(1)
)
tutorial_dependency_text <- paste(
  paste0(tutorial_dependency_packages, " {", tutorial_dependency_versions, "}"),
  collapse = "; "
)
tutorial_bioc_version <- if (requireNamespace("BiocManager", quietly = TRUE)) {
  as.character(BiocManager::version())
} else {
  "not installed"
}
```

<div class="tutorial-title-block">
<h1 class="title">Comparative ranked GSEA waterfall, volcano, scatter-style plots</h1>
<div class="tutorial-subtitle">R functions for plotting ranked GSEA GO terms and categorizing them by biological function, as well as comparative cross-contrast GSEA NES scatterplots.</div>

<div class="tutorial-meta-block">
<div><span>Authors:</span> Zoheb Khan</div>
<div><span>Modified:</span> January 24, 2026</div>
<div><span>Compiled:</span> May 9, 2026</div>
<div><span>Environment:</span> `r R.version.string`; Bioconductor `r tutorial_bioc_version`</div>
<div><span>Dependencies:</span> `r tutorial_dependency_text`</div>
<div><span>Source:</span> repository root (`.`)</div>
</div>
</div>

# Overview

This workflow provides sourceable R plotting functions for precomputed Gene Set Enrichment Analysis (GSEA) results. It does not run DESeq2, build ranked gene lists, or calculate enrichment. Those steps should stay in the analysis pipeline that produced the GSEA result table.

Rendered tutorial: `https://zohebkhan1.github.io/gsea-waterfall/`

The plotting functions generate four figure types:

1. ranked positive and negative NES waterfall plots
2. a symmetric GSEA volcano plot
3. directional half-volcano plots
4. comparative NES scatterplots across two GSEA contrasts

The worked example uses GO biological-process GSEA results from GSE122380 cardiomyocyte differentiation contrasts. The example input CSVs are deliberately small and contain only the columns needed for plotting.

# Get The Functions

Download only the plotting functions from GitHub:

```r
# set the raw github function path
base_url <- "https://raw.githubusercontent.com/ZohebKhan1/gsea-waterfall/main/functions"

download.file(paste0(base_url, "/gsea_waterfall_plot.R"), "gsea_waterfall_plot.R")
download.file(paste0(base_url, "/gsea_volcano_plot.R"), "gsea_volcano_plot.R")
download.file(paste0(base_url, "/gsea_half_volcano_plot.R"), "gsea_half_volcano_plot.R")
download.file(paste0(base_url, "/gsea_nes_scatter_plot.R"), "gsea_nes_scatter_plot.R")

source("gsea_waterfall_plot.R")
source("gsea_volcano_plot.R")
source("gsea_half_volcano_plot.R")
source("gsea_nes_scatter_plot.R")
```

Cloning the repository downloads the public tutorial source, plotting functions, minimal example CSVs, rendered assets, and GitHub Pages site. Private inputs, full result tables, reference folders, internal scripts, and local agent files are excluded by `.gitignore`.

# Input Table

Each plot function expects a precomputed GSEA result table. Use this header:

| Column | Required | Meaning |
|:---|:---|:---|
| `go_description` | yes | GO term name |
| `NES` | yes | normalized enrichment score |
| `pval` | yes | nominal GSEA p-value |
| `padj` | yes | adjusted p-value |

These four columns are the only columns users need for the plotting workflow shown here.

# Waterfall Plots

Use `plot_gsea_waterfall()` for a ranked view of enriched GO terms in one NES direction. Terms can be ranked by `NES`, `padj`, or `pvalue`. Labels can be selected by exact term name, GO ID when available, plotted rank, or automatic rank-based selection.

`term_groups` and `label_terms` are intentionally separate. `term_groups` defines broad biological categories for point colors; `label_terms` defines the smaller set of GO terms that should receive text labels. By default, category matching uses exact GO IDs and text seeds. Set `include_descendants = TRUE` only when category seeds include GO IDs and GO offspring should be included through `GO.db`.

## Positive NES waterfall

```r
# define color groups
positive_term_groups <- list(
  "Ion channel" = c(
    "Calcium", "Sodium", "Potassium", "Ion", "Voltage", "Channel",
    "electrical", "Action potential", "transmembrane", "conduction"
  ),
  "Metabolism" = c(
    "Metabolic", "Metabolism", "Electron transport chain", "Mitochondria",
    "Mitochondrial", "Oxidative", "Phosphorylation", "ATP", "lipid",
    "biosynthetic", "respiration"
  ),
  "Muscle contraction" = c(
    "Muscle contraction", "Muscle cell differentiation", "Striated",
    "Sarcomere", "Myofibril", "Actin", "Muscle"
  ),
  "Heart development" = c(
    "atrial", "ventricular", "heart", "cardiac"
  )
)

# define labeled terms
positive_label_terms <- c(
  "myofibril assembly",
  "cardiac muscle contraction",
  "cardiac muscle cell action potential",
  "calcium-mediated signaling",
  "electron transport chain",
  "muscle cell differentiation",
  "ATP metabolic process",
  "cardiac cell development",
  "cardiac chamber morphogenesis",
  "muscle contraction",
  "cardiac conduction",
  "cardiac muscle cell contraction",
  "heart growth",
  "cardiac atrium development",
  "heart trabecula morphogenesis"
)

# plot ranked positive terms
positive_waterfall <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = "positive",
  top_n = 100,
  rank_by = "NES",
  x_label = "Ranked GO terms by NES for Cardiomyocyte vs. Mesoderm",
  label_n = 10,
  label_terms = positive_label_terms,
  term_groups = positive_term_groups,
  category_n = 4
)
```

```{r positive-waterfall-figure, echo=FALSE, results='asis', out.width='80%', fig.align='center'}
cat('<div id="fig-positive-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics("assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_positive.svg")
```

<div class="tutorial-figure-caption"><strong>Figure 1.</strong> Positive NES GO biological-process waterfall for Cardiomyocyte vs. Mesoderm. Terms are ranked by NES, colored by manually defined cardiac-relevant GO term groups, and labeled with selected high-priority terms.</div>

## Negative NES waterfall

```r
# define negative color groups
negative_term_groups <- list(
  "Ribosomal" = c(
    "ribosome", "ribosomal", "spliceosomal", "rRNA", "RNA"
  ),
  "Mitosis" = c(
    "Mitosis", "Meiosis", "Nuclear division", "cell cycle", "telomere",
    "spindle", "Mitotic", "Meiotic"
  ),
  "DNA replication" = c(
    "chromosome", "DNA", "base", "repair", "DNA-"
  )
)

# set negative group colors
negative_term_group_colors <- c(
  "Ribosomal" = "#377EB8",
  "Mitosis" = "#984EA3",
  "DNA replication" = "#E41A1C"
)

negative_waterfall <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = "negative",
  top_n = 100,
  rank_by = "NES",
  x_label = "Ranked GO terms by NES for Cardiomyocyte vs. Mesoderm",
  label_n = 10,
  term_groups = negative_term_groups,
  term_group_colors = negative_term_group_colors
)
```

```{r negative-waterfall-figure, echo=FALSE, results='asis', out.width='80%', fig.align='center'}
cat('<div id="fig-negative-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics("assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_negative.svg")
```

<div class="tutorial-figure-caption"><strong>Figure 2.</strong> Negative NES GO biological-process waterfall for the same contrast. The selected groups highlight ribosomal, mitotic, and DNA-replication programs.</div>

# Volcano Plots

Use `plot_gsea_volcano()` when positive and negative enrichment should be inspected in one symmetric view. Use `plot_gsea_half_volcano()` when one NES direction needs a larger directional panel.

## Symmetric volcano

```r
# define volcano labels
positive_volcano_label_terms <- c(
  "cardiac muscle contraction",
  "muscle cell differentiation",
  "cardiac chamber morphogenesis",
  "heart growth",
  "heart valve development"
)

negative_volcano_label_terms <- c(
  "mitotic nuclear division",
  "DNA replication initiation",
  "mitotic G2/M transition checkpoint",
  "centriole assembly",
  "ribosome assembly"
)

symmetric_volcano <- plot_gsea_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  label_n = 10,
  label_terms = c(positive_volcano_label_terms, negative_volcano_label_terms),
  contrast_label = "Cardiomyocyte vs. Mesoderm",
  label_words_per_line = 3
)
```

```{r symmetric-volcano-figure, echo=FALSE, results='asis', out.width='58%', fig.align='center'}
cat('<div id="fig-symmetric-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics("assets/figures/GSE122380_gsea_volcano_cardiomyocyte_vs_mesoderm.svg")
```

<div class="tutorial-figure-caption"><strong>Figure 3.</strong> Symmetric GO biological-process GSEA volcano for Cardiomyocyte vs. Mesoderm. NES is shown on the x-axis, adjusted p-value on the y-axis, and point color marks significant positive, significant negative, or non-significant enrichment.</div>

## Directional half-volcano plots

```r
positive_half_volcano_label_terms <- c(
  "myofibril assembly",
  "cardiac muscle contraction",
  "cardiac muscle cell action potential",
  "electron transport chain",
  "muscle cell differentiation",
  "cardiac cell development",
  "cardiac chamber morphogenesis",
  "muscle contraction",
  "cardiac conduction",
  "heart growth",
  "heart trabecula morphogenesis"
)

positive_half_volcano <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = "positive",
  label_n = 10,
  label_terms = positive_half_volcano_label_terms,
  term_groups = positive_term_groups,
  category_n = 4,
  contrast_label = "Cardiomyocyte vs. Mesoderm"
)
```

```{r positive-half-volcano-figure, echo=FALSE, results='asis', out.width='72%', fig.align='center'}
cat('<div id="fig-positive-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics("assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_positive.svg")
```

<div class="tutorial-figure-caption"><strong>Figure 4.</strong> Positive NES half-volcano for Cardiomyocyte vs. Mesoderm. The enlarged directional view uses the same labeled positive terms as the positive NES waterfall plot.</div>

```r
# define negative labels
negative_label_terms <- c(
  "DNA-templated DNA replication",
  "DNA replication",
  "cell cycle checkpoint signaling",
  "mitotic cell cycle checkpoint signaling",
  "ribosome biogenesis",
  "recombinational repair",
  "DNA replication initiation",
  "cell cycle DNA replication",
  "spindle organization"
)

negative_half_volcano <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = "negative",
  label_n = 10,
  label_terms = negative_label_terms,
  term_groups = negative_term_groups,
  term_group_colors = negative_term_group_colors,
  contrast_label = "Cardiomyocyte vs. Mesoderm"
)
```

```{r negative-half-volcano-figure, echo=FALSE, results='asis', out.width='72%', fig.align='center'}
cat('<div id="fig-negative-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics("assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_negative.svg")
```

<div class="tutorial-figure-caption"><strong>Figure 5.</strong> Negative NES half-volcano for Cardiomyocyte vs. Mesoderm. The selected terms emphasize DNA replication, checkpoint, and ribosome-associated programs.</div>

# Comparative NES Scatterplots

Use `plot_gsea_nes_scatter()` to compare the same GO terms across two GSEA result tables. The scatterplots below compare late and early cardiomyocyte differentiation intervals.

```r
# define all-quadrant labels
all_quadrants_scatter_label_terms <- c(
  "mesenchyme development",
  "mesoderm development",
  "mesodermal cell differentiation",
  "myofibril assembly",
  "heart process",
  "oxidative phosphorylation",
  "regulation of nuclear division"
)
```

## All quadrants

```r
# plot all NES quadrants
all_quadrants_scatter <- plot_gsea_nes_scatter(
  gsea_x = gsea_day9_vs_day6,
  gsea_y = gsea_day3_vs_day1,
  x_label = "Day 9 vs. Day 6 NES",
  y_label = "Day 3 vs. Day 1 NES",
  color_by = "significance",
  quadrant = "all",
  x_name = "Day 9 vs. Day 6",
  y_name = "Day 3 vs. Day 1",
  label_terms = all_quadrants_scatter_label_terms,
  include_nonsignificant = TRUE,
  equal_axis_limits = TRUE
)
```

```{r all-quadrants-scatter-figure, echo=FALSE, results='asis', out.width='100%', fig.align='center'}
cat('<div id="fig-all-quadrants-scatter" class="figure-anchor"></div>\n')
knitr::include_graphics("assets/figures/GSE122380_gsea_scatter_day9_vs_day6_x_day3_vs_day1_all_quadrants.svg")
```

<div class="tutorial-figure-caption"><strong>Figure 6.</strong> Comparative GO biological-process NES scatterplot for Day 9 vs. Day 6 and Day 3 vs. Day 1. The fitted line summarizes the NES relationship across plotted GO terms, while point color indicates whether each GO term is significant in one, both, or neither contrast.</div>

# Session Info

```{r session-info}
sessionInfo()
```

# References

<ol class="reference-list">
<li>Subramanian A, Tamayo P, Mootha VK, et al. Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. <em>PNAS</em>. 2005. <a href="https://www.pnas.org/doi/10.1073/pnas.0506580102">https://www.pnas.org/doi/10.1073/pnas.0506580102</a></li>
<li>Korotkevich G, Sukhov V, Budin N, Shpak B, Artyomov MN, Sergushichev A. Fast gene set enrichment analysis. <a href="https://www.biorxiv.org/content/10.1101/060012v3">https://www.biorxiv.org/content/10.1101/060012v3</a></li>
<li>Bioconductor `fgsea`: Fast Gene Set Enrichment Analysis. <a href="https://bioconductor.org/packages/fgsea/">https://bioconductor.org/packages/fgsea/</a></li>
<li>The Gene Ontology Consortium. The Gene Ontology resource. <a href="https://geneontology.org/">https://geneontology.org/</a></li>
</ol>
