---
title: "GSEA waterfall, volcano, and comparative NES plots"
description: "Standalone R functions for plotting precomputed GSEA results."
site: bookdown::bookdown_site
documentclass: book
---

```{r tutorial-options, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
```

<div class="tutorial-title-block">
<h1 class="title">GSEA waterfall, volcano, and comparative NES plots</h1>
<div class="tutorial-subtitle">Standalone R functions for visualizing ranked GO biological-process enrichment results.</div>

<div class="tutorial-meta-block">
<div><span>Author:</span> Zoheb Khan</div>
<div><span>Modified:</span> 2026-08-02</div>
<div><span>Compiled:</span> 2026-08-02</div>
<div><span>Environment:</span> R version 4.6.1 (2026-06-24)</div>
<div><span>Bioconductor:</span> 3.23</div>
<div><span>Example:</span> GSE122380 GO biological-process GSEA contrasts</div>
<div><span>Source:</span> <a href="https://github.com/ZohebKhan1/gsea-waterfall">GitHub repository</a></div>
</div>
</div>

# Overview

`gsea-waterfall` visualizes **precomputed** Gene Set Enrichment Analysis
(GSEA) results as ranked NES waterfalls, volcano plots, and cross-contrast
comparative NES scatterplots. The functions return `ggplot` objects. They are a
visualization layer applied after GSEA has been performed and do not change
supplied statistics or write files.

The examples use fixed GO biological-process results from the GSE122380
cardiomyocyte iPSC time-course bulk RNA-seq contrasts. Plot controls select,
group, and label displayed rows; significance classes use the explicit
adjusted-p-value cutoff without altering the supplied statistics.

The code below is the user-facing workflow. The exact export settings used to
produce the bundled SVG files are maintained in `R/01_generate_gsea_figures.R`;
the tutorial keeps those export-only styling details out of the main workflow.
The figures below are the bundled examples produced by that export script, so
their labels and category seeds may be more specific than the compact calls
shown here.

## What is GSEA?

Gene Set Enrichment Analysis evaluates whether the genes in a predefined gene
set are concentrated toward the top or bottom of a ranked gene list. Rather
than testing genes one at a time, GSEA summarizes coordinated changes across
biological processes or pathways. A typical result table contains one row per
gene set with a term description, normalized enrichment score (NES), nominal
p-value, and adjusted p-value.

GSEA can be performed with any appropriate tool, including `fgsea`,
`clusterProfiler`, or another package. This repository accepts completed GSEA
result tables and does not calculate enrichment statistics or perform
multiple-testing correction. At minimum, a waterfall needs a term description
and NES. The shared reader also validates nominal and adjusted p-values so the
same standardized table can support volcano and comparative plots.

NES is the primary direction and ranking measure: positive and negative values
indicate enrichment on opposite sides of the ranked gene list. Adjusted
p-values describe statistical significance and are used by the volcano and
comparative plots for coloring and display decisions. NES and adjusted
p-values answer different questions and should not be interpreted
interchangeably. The plots use these fields differently: waterfalls rank by
NES or a p-value, volcanoes show NES against adjusted p-value, half-volcanoes
can show nominal or adjusted p-value, and comparative scatterplots use NES with
adjusted-p-value significance classes.

## Why use GSEA waterfall plots?

GSEA results are often presented as a running-score enrichment plot when one
pathway is the focus, or as a dotplot/barplot-style figure in which NES or
adjusted p-value is encoded on the x-axis and other GSEA metrics are shown by
point size or color. These compact plots are useful for a short list of terms,
but they become crowded when many results are displayed and commonly focus on
only the top 10–20 terms.

The top-ranked terms can also be functionally redundant: related GO parent and
child terms may appear next to one another because they represent overlapping
gene sets. A waterfall plot does not remove this redundancy automatically, but
it provides a broader view of the ranked enrichment landscape, often allowing
100–200 terms to be inspected rather than only the first few rows.

The waterfall functions also allow displayed GO terms to be grouped using
caller-supplied keyword or identifier categories. In the cardiomyocyte iPSC
time-course bulk RNA-seq example, these annotations highlight processes such
as cardiac development, metabolism, ribosomal activity, mitosis, and DNA
replication. They are visualization annotations, not additional statistical
tests; terms remain neutral by default unless `term_groups` is supplied.

The initial idea for the waterfall presentation was informed by pathway-level
figures in [Ciceri et al., *Nature*, 2024](https://doi.org/10.1038/s41586-023-06984-8),
[Xu et al., *Nature Cell Biology*, 2025](https://doi.org/10.1038/s41556-025-01751-5),
[Vuong et al., *Science*, 2026](https://doi.org/10.1126/science.aea1259), and
[Risgaard et al., *Science*, 2026](https://doi.org/10.1126/science.aea1549).

# Install and source

Install the plotting dependencies once:

```r
install.packages(c('ggplot2', 'ggrepel', 'RColorBrewer', 'scales'))
```

Download the shared utilities first, then the plot modules:

```r
base_url <- 'https://raw.githubusercontent.com/ZohebKhan1/gsea-waterfall/main/functions'

download.file(paste0(base_url, '/gsea_plot_utils.R'), 'gsea_plot_utils.R')
download.file(paste0(base_url, '/gsea_waterfall_plot.R'), 'gsea_waterfall_plot.R')
download.file(paste0(base_url, '/gsea_volcano_plot.R'), 'gsea_volcano_plot.R')
download.file(paste0(base_url, '/gsea_half_volcano_plot.R'), 'gsea_half_volcano_plot.R')
download.file(paste0(base_url, '/gsea_nes_scatter_plot.R'), 'gsea_nes_scatter_plot.R')

source('gsea_plot_utils.R')
source('gsea_waterfall_plot.R')
source('gsea_volcano_plot.R')
source('gsea_half_volcano_plot.R')
source('gsea_nes_scatter_plot.R')
```

# Input data

Each table represents one GSEA contrast with one row per unique term. The default columns are:

| Column | Required | Meaning |
|:---|:---|:---|
| `go_description` | yes | GO term description and fallback matching key |
| `go_term_id` | no | stable matching key used automatically when present |
| `NES` | yes | normalized enrichment score |
| `pval` | yes | nominal GSEA p-value in `[0, 1]` |
| `padj` | yes | adjusted p-value in `[0, 1]` |

Required columns must be complete, and NES values must be finite. Key precedence is explicit `id_col`, then `go_term_id` when present, otherwise `go_description`. Comparative scatterplots use the exact shared-key intersection and do not impute absent terms.

If your result table uses different column names, map them when reading it:

```r
fgsea_results <- read_gsea_result_csv(
  'fgsea_results.csv',
  term_col = 'pathway',
  nes_col = 'NES',
  pvalue_col = 'pval',
  padj_col = 'padj'
)
```

For comparative plots, use `id_col` with a stable term identifier whenever one
is available. Without stable IDs, term descriptions must match exactly across
tables.

Load the bundled examples from the repository root:

```r
gsea_cardiomyocyte_vs_mesoderm <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_cardiomyocyte_vs_mesoderm.csv'
)
gsea_day9_vs_day6 <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_day9_vs_day6.csv'
)
gsea_day3_vs_day1 <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_day3_vs_day1.csv'
)

print(head(gsea_cardiomyocyte_vs_mesoderm, 5L))
```

# Ranked GSEA waterfall plots

`plot_gsea_waterfall()` selects one NES direction, ranks by `NES`, `padj`, `pvalue`, or `pval`, and retains `top_n` terms. `term_groups` controls colors and `label_terms` controls labels. Matches are case-insensitive against term descriptions or IDs.

## Positive NES waterfall

```r
positive_term_groups <- list(
  'Cardiac development' = c('cardiac', 'heart', 'muscle'),
  'Metabolism' = c('mitochondrial', 'metabolic', 'oxidative')
)

positive_waterfall_plot <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'positive',
  top_n = 100L,
  rank_by = 'NES',
  label_n = 12L,
  term_groups = positive_term_groups
)
```

`direction` chooses the NES side to display. `top_n` is applied after that
filter and after ranking. Terms are neutral unless `term_groups` is supplied;
each seed is matched case-insensitively against term IDs or descriptions.

```{r positive-waterfall-figure, results='asis', out.width='80%', fig.alt='Positive NES waterfall for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-positive-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_positive.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 1.</strong> Positive NES terms ranked by NES, with selected term groups and labels.</div>

## Negative NES waterfall

```r
negative_waterfall_plot <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'negative',
  top_n = 100L,
  rank_by = 'NES',
  label_n = 10L
)
```

Use `rank_by = 'padj'` or `rank_by = 'pvalue'` when statistical significance,
rather than NES magnitude, should determine the displayed order. `label_n`
selects labels automatically; use `label_terms` when specific terms should be
shown.

```{r negative-waterfall-figure, results='asis', out.width='80%', fig.alt='Negative NES waterfall for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-negative-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_negative.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 2.</strong> Bundled negative NES example with caller-defined ribosomal, mitotic, and DNA-replication term groups.</div>

# Volcano plots

Use `plot_gsea_volcano()` for both NES directions in one panel and `plot_gsea_half_volcano()` for a directional view. All examples use the established adjusted p-value cutoff of 0.05.

## Symmetric volcano

```r
symmetric_volcano_plot <- plot_gsea_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  padj_cutoff = 0.05,
  label_n = 10L,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

`padj_cutoff` controls significance coloring and the displayed counts. It does
not remove rows from the plot. Use `label_terms` instead of `label_n` when you
want to label exact terms.

```{r symmetric-volcano-figure, results='asis', out.width='58%', fig.alt='Symmetric GSEA volcano for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-symmetric-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_volcano_cardiomyocyte_vs_mesoderm.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 3.</strong> Symmetric NES volcano colored by adjusted-p-value significance and direction.</div>

## Directional half-volcano plots

```r
positive_half_volcano_plot <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'positive',
  p_col = 'padj',
  padj_cutoff = 0.05,
  label_n = 10L,
  term_groups = positive_term_groups,
  inner_nes_limit = 1,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

`inner_nes_limit = 1` keeps terms with positive NES at least 1 in magnitude.
Set `p_col = 'pvalue'` to show nominal p-values instead; significance coloring
continues to use `padj_cutoff`.

```{r positive-half-volcano-figure, results='asis', out.width='72%', fig.alt='Positive NES half-volcano for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-positive-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_positive.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 4.</strong> Bundled positive NES half-volcano with caller-defined term groups and labels.</div>

```r
negative_half_volcano_plot <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'negative',
  p_col = 'padj',
  padj_cutoff = 0.05,
  label_n = 10L,
  inner_nes_limit = 1,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

```{r negative-half-volcano-figure, results='asis', out.width='72%', fig.alt='Negative NES half-volcano for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-negative-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_negative.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 5.</strong> Bundled negative NES half-volcano with caller-defined term groups and labels.</div>

# Comparative NES scatterplot

`plot_gsea_nes_scatter()` compares exact shared term keys. By default it keeps
terms significant in at least one contrast and does not draw a fitted line. The
line is descriptive when requested and does not alter enrichment statistics.

```r
nes_scatter_plot <- plot_gsea_nes_scatter(
  gsea_x = gsea_day9_vs_day6,
  gsea_y = gsea_day3_vs_day1,
  x_label = 'Day 9 vs. Day 6 NES',
  y_label = 'Day 3 vs. Day 1 NES',
  color_by = 'significance',
  quadrant = 'all',
  x_name = 'Day 9 vs. Day 6',
  y_name = 'Day 3 vs. Day 1',
  label_n_x_only = 3L,
  label_n_y_only = 3L,
  label_n_both = 3L,
  padj_cutoff = 0.05,
  include_nonsignificant = TRUE,
  equal_axis_limits = TRUE,
  show_fit_line = TRUE
)
```

```{r all-quadrants-scatter-figure, results='asis', out.width='100%', fig.alt='Comparative NES scatterplot for Day 9 versus Day 6 and Day 3 versus Day 1'}
cat('<div id="fig-all-quadrants-scatter" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_scatter_day9_vs_day6_x_day3_vs_day1_all_quadrants.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 6.</strong> Shared terms colored by significance in one, both, or neither contrast.</div>

Set `include_nonsignificant = FALSE` to focus on terms significant in at least
one contrast. Use `quadrant = 'q1'` or `quadrant = 'q3'` to focus on concordant
positive or negative NES values. `equal_axis_limits = TRUE` makes the two NES
axes directly comparable.

# API summary

| Function | Contract |
|:---|:---|
| `read_gsea_result_csv()` | reads and validates one precomputed CSV without filtering rows |
| `plot_gsea_waterfall()` | returns a ranked one-direction ggplot |
| `plot_gsea_volcano()` | returns a symmetric NES-versus-adjusted-p-value ggplot |
| `plot_gsea_half_volcano()` | returns a directional ggplot after the explicit NES boundary |
| `plot_gsea_nes_scatter()` | returns a ggplot from the exact shared term-key intersection |

Source documentation defines the complete argument contracts, including the common `term_col`, `nes_col`, `pvalue_col`, `padj_col`, and optional `id_col` arguments.

# References

<ol class="reference-list">
<li>Subramanian A, Tamayo P, Mootha VK, et al. Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. <em>PNAS</em>. 2005. <a href="https://www.pnas.org/doi/10.1073/pnas.0506580102">https://www.pnas.org/doi/10.1073/pnas.0506580102</a></li>
<li>Ciceri G, Baggiolini A, Cho HS, et al. An epigenetic barrier sets the timing of human neuronal maturation. <em>Nature</em>. 2024;626:881-890. <a href="https://doi.org/10.1038/s41586-023-06984-8">https://doi.org/10.1038/s41586-023-06984-8</a></li>
<li>Xu N, Cho HS, Hackland JOS, et al. Genome-wide CRISPR screen identifies Menin and SUZ12 as regulators of human developmental timing. <em>Nature Cell Biology</em>. 2025;27:1411-1421. <a href="https://doi.org/10.1038/s41556-025-01751-5">https://doi.org/10.1038/s41556-025-01751-5</a></li>
<li>Vuong CK, Weber A, Seong P, et al. A single-cell multiomic analysis identifies molecular and gene-regulatory mechanisms dysregulated in developing Down syndrome neocortex. <em>Science</em>. 2026;392:eaea1259. <a href="https://doi.org/10.1126/science.aea1259">https://doi.org/10.1126/science.aea1259</a></li>
<li>Risgaard RD, et al. Molecular and cellular processes disrupted in the early postnatal Down syndrome prefrontal cortex. <em>Science</em>. 2026;392:eaea1549. <a href="https://doi.org/10.1126/science.aea1549">https://doi.org/10.1126/science.aea1549</a></li>
</ol>
