---
title: "GSEA waterfall, volcano, and comparative NES plots"
description: "Standalone R functions for plotting precomputed GSEA results."
site: bookdown::bookdown_site
documentclass: book
---

```{r tutorial-options, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
```

<div class="tutorial-title-block">
<h1 class="title">GSEA waterfall, volcano, and comparative NES plots</h1>
<div class="tutorial-subtitle">Standalone R functions for visualizing ranked GO biological-process enrichment results.</div>

<div class="tutorial-meta-block">
<div><span>Author:</span> Zoheb Khan</div>
<div><span>Modified:</span> 2026-08-01</div>
<div><span>Compiled:</span> 2026-08-01</div>
<div><span>Environment:</span> R version 4.6.1 (2026-06-24); Bioconductor 3.23</div>
<div><span>Example:</span> GSE122380 GO biological-process GSEA contrasts</div>
<div><span>Source:</span> <a href="https://github.com/ZohebKhan1/gsea-waterfall">GitHub repository</a></div>
</div>
</div>

# Overview

`gsea-waterfall` visualizes **precomputed** Gene Set Enrichment Analysis (GSEA) results as ranked NES waterfalls, symmetric or directional volcano plots, and cross-contrast NES scatterplots. The functions return ggplot objects without running GSEA, changing supplied statistics, or writing files.

The examples use fixed GO biological-process results from GSE122380 cardiomyocyte differentiation contrasts. Plot controls select, group, and label displayed rows; significance classes use the explicit adjusted-p-value cutoff without altering the supplied statistics.

The waterfall layout is inspired by pathway-level figures in [Ciceri, *Nature*, 2024](https://doi.org/10.1038/s41586-023-06984-8), [Xu, *Nature Cell Biology*, 2025](https://doi.org/10.1038/s41556-025-01751-5), [Vuong, *Science*, 2026](https://doi.org/10.1126/science.aea1259), and [Risgaard, *Science*, 2026](https://doi.org/10.1126/science.aea1549).

# Install and source

Install the plotting dependencies once:

```r
utils::install.packages(c('ggplot2', 'ggrepel', 'RColorBrewer', 'scales'))
```

Download the shared utilities first, then the plot modules:

```r
base_url <- 'https://raw.githubusercontent.com/ZohebKhan1/gsea-waterfall/main/functions'

utils::download.file(paste0(base_url, '/gsea_plot_utils.R'), 'gsea_plot_utils.R')
utils::download.file(paste0(base_url, '/gsea_waterfall_plot.R'), 'gsea_waterfall_plot.R')
utils::download.file(paste0(base_url, '/gsea_volcano_plot.R'), 'gsea_volcano_plot.R')
utils::download.file(paste0(base_url, '/gsea_half_volcano_plot.R'), 'gsea_half_volcano_plot.R')
utils::download.file(paste0(base_url, '/gsea_nes_scatter_plot.R'), 'gsea_nes_scatter_plot.R')

source('gsea_plot_utils.R')
source('gsea_waterfall_plot.R')
source('gsea_volcano_plot.R')
source('gsea_half_volcano_plot.R')
source('gsea_nes_scatter_plot.R')
```

# Input contract

Each table represents one GSEA contrast with one row per unique term. The default columns are:

| Column | Required | Meaning |
|:---|:---|:---|
| `go_description` | yes | GO term description and fallback matching key |
| `go_term_id` | no | stable matching key used automatically when present |
| `NES` | yes | normalized enrichment score |
| `pval` | yes | nominal GSEA p-value in `[0, 1]` |
| `padj` | yes | adjusted p-value in `[0, 1]` |

Required columns must be complete, and NES values must be finite. Key precedence is explicit `id_col`, then `go_term_id` when present, otherwise `go_description`. Comparative scatterplots use the exact shared-key intersection and do not impute absent terms.

Load the bundled examples from the repository root:

```r
gsea_cardiomyocyte_vs_mesoderm <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_cardiomyocyte_vs_mesoderm.csv'
)
gsea_day9_vs_day6 <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_day9_vs_day6.csv'
)
gsea_day3_vs_day1 <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_day3_vs_day1.csv'
)
```

# Ranked waterfall plots

`plot_gsea_waterfall()` selects one NES direction, ranks by `NES`, `padj`, `pvalue`, or `pval`, and retains `top_n` terms. `term_groups` controls colors and `label_terms` controls labels. Matches are case-insensitive against term descriptions or IDs.

## Positive NES waterfall

```r
positive_term_groups <- list(
  'Ion channel' = c(
    'Calcium', 'Sodium', 'Potassium', 'Ion', 'Voltage', 'Channel',
    'electrical', 'Action potential', 'transmembrane', 'conduction'
  ),
  'Metabolism' = c(
    'Metabolic', 'Metabolism', 'Electron transport chain', 'Mitochondria',
    'Mitochondrial', 'Oxidative', 'Phosphorylation', 'ATP', 'lipid',
    'biosynthetic', 'respiration'
  ),
  'Muscle contraction' = c(
    'Muscle contraction', 'Muscle cell differentiation', 'Striated',
    'Sarcomere', 'Myofibril', 'Actin', 'Muscle'
  ),
  'Heart development' = c('atrial', 'ventricular', 'heart', 'cardiac')
)

positive_waterfall_label_terms <- c(
  'myofibril assembly',
  'cardiac muscle contraction',
  'cardiac muscle cell action potential',
  'calcium-mediated signaling',
  'electron transport chain',
  'muscle cell differentiation',
  'ATP metabolic process',
  'cardiac cell development',
  'cardiac chamber morphogenesis',
  'muscle contraction',
  'cardiac conduction',
  'cardiac muscle cell contraction',
  'heart growth',
  'cardiac atrium development',
  'heart trabecula morphogenesis'
)

positive_waterfall_plot <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'positive',
  top_n = 100L,
  rank_by = 'NES',
  x_label = 'Ranked GO terms by NES for Cardiomyocyte vs. Mesoderm',
  label_terms = positive_waterfall_label_terms,
  term_groups = positive_term_groups
)
```

```{r positive-waterfall-figure, results='asis', out.width='80%', fig.alt='Positive NES waterfall for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-positive-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_positive.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 1.</strong> Positive NES terms ranked by NES, with selected term groups and labels.</div>

## Negative NES waterfall

```r
negative_term_groups <- list(
  'Ribosomal' = c('ribosome', 'ribosomal', 'spliceosomal', 'rRNA', 'RNA'),
  'Mitosis' = c(
    'Mitosis', 'Meiosis', 'Nuclear division', 'cell cycle', 'telomere',
    'spindle', 'Mitotic', 'Meiotic'
  ),
  'DNA replication' = c('chromosome', 'DNA', 'base', 'repair', 'DNA-')
)

negative_term_group_colors <- c(
  'Ribosomal' = '#377EB8',
  'Mitosis' = '#984EA3',
  'DNA replication' = '#E41A1C'
)

negative_waterfall_plot <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'negative',
  top_n = 100L,
  rank_by = 'NES',
  x_label = 'Ranked GO terms by NES for Cardiomyocyte vs. Mesoderm',
  label_n = 10L,
  term_groups = negative_term_groups,
  term_group_colors = negative_term_group_colors
)
```

```{r negative-waterfall-figure, results='asis', out.width='80%', fig.alt='Negative NES waterfall for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-negative-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_negative.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 2.</strong> Negative NES terms highlighting ribosomal, mitotic, and DNA-replication term groups.</div>

# Volcano plots

Use `plot_gsea_volcano()` for both NES directions in one panel and `plot_gsea_half_volcano()` for a directional view. All examples use the established adjusted p-value cutoff of 0.05.

## Symmetric volcano

```r
volcano_label_terms <- c(
  'cardiac muscle contraction',
  'muscle cell differentiation',
  'cardiac chamber morphogenesis',
  'heart growth',
  'heart valve development',
  'mitotic nuclear division',
  'DNA replication initiation',
  'mitotic G2/M transition checkpoint',
  'centriole assembly',
  'ribosome assembly'
)

symmetric_volcano_plot <- plot_gsea_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  padj_cutoff = 0.05,
  label_terms = volcano_label_terms,
  contrast_label = 'Cardiomyocyte vs. Mesoderm',
  label_words_per_line = 3L
)
```

```{r symmetric-volcano-figure, results='asis', out.width='58%', fig.alt='Symmetric GSEA volcano for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-symmetric-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_volcano_cardiomyocyte_vs_mesoderm.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 3.</strong> Symmetric NES volcano colored by adjusted-p-value significance and direction.</div>

## Directional half-volcano plots

```r
positive_half_volcano_label_terms <- c(
  'myofibril assembly',
  'cardiac muscle contraction',
  'cardiac muscle cell action potential',
  'electron transport chain',
  'muscle cell differentiation',
  'cardiac cell development',
  'cardiac chamber morphogenesis',
  'muscle contraction',
  'cardiac conduction',
  'heart growth',
  'heart trabecula morphogenesis'
)

positive_half_volcano_plot <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'positive',
  p_col = 'padj',
  padj_cutoff = 0.05,
  label_terms = positive_half_volcano_label_terms,
  term_groups = positive_term_groups,
  inner_nes_limit = 1,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

```{r positive-half-volcano-figure, results='asis', out.width='72%', fig.alt='Positive NES half-volcano for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-positive-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_positive.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 4.</strong> Positive NES half-volcano using the waterfall plot's term groups.</div>

```r
negative_half_volcano_label_terms <- c(
  'DNA-templated DNA replication',
  'DNA replication',
  'cell cycle checkpoint signaling',
  'mitotic cell cycle checkpoint signaling',
  'ribosome biogenesis',
  'recombinational repair',
  'DNA replication initiation',
  'cell cycle DNA replication',
  'spindle organization'
)

negative_half_volcano_plot <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'negative',
  p_col = 'padj',
  padj_cutoff = 0.05,
  label_terms = negative_half_volcano_label_terms,
  term_groups = negative_term_groups,
  term_group_colors = negative_term_group_colors,
  inner_nes_limit = 1,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

```{r negative-half-volcano-figure, results='asis', out.width='72%', fig.alt='Negative NES half-volcano for Cardiomyocyte versus Mesoderm'}
cat('<div id="fig-negative-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_negative.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 5.</strong> Negative NES half-volcano emphasizing DNA replication, checkpoint, and ribosome-associated terms.</div>

# Comparative NES scatterplot

`plot_gsea_nes_scatter()` compares exact shared term keys. The fitted line is descriptive and does not alter enrichment statistics.

```r
nes_scatter_label_terms <- c(
  'mesenchyme development',
  'mesoderm development',
  'mesodermal cell differentiation',
  'myofibril assembly',
  'heart process',
  'oxidative phosphorylation',
  'regulation of nuclear division'
)

nes_scatter_plot <- plot_gsea_nes_scatter(
  gsea_x = gsea_day9_vs_day6,
  gsea_y = gsea_day3_vs_day1,
  x_label = 'Day 9 vs. Day 6 NES',
  y_label = 'Day 3 vs. Day 1 NES',
  color_by = 'significance',
  quadrant = 'all',
  x_name = 'Day 9 vs. Day 6',
  y_name = 'Day 3 vs. Day 1',
  label_terms = nes_scatter_label_terms,
  padj_cutoff = 0.05,
  include_nonsignificant = TRUE,
  equal_axis_limits = TRUE,
  show_fit_line = TRUE
)
```

```{r all-quadrants-scatter-figure, results='asis', out.width='100%', fig.alt='Comparative NES scatterplot for Day 9 versus Day 6 and Day 3 versus Day 1'}
cat('<div id="fig-all-quadrants-scatter" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_scatter_day9_vs_day6_x_day3_vs_day1_all_quadrants.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 6.</strong> Shared terms colored by significance in one, both, or neither contrast.</div>

# API summary

| Function | Contract |
|:---|:---|
| `read_gsea_result_csv()` | reads and validates one precomputed CSV without filtering rows |
| `plot_gsea_waterfall()` | returns a ranked one-direction ggplot |
| `plot_gsea_volcano()` | returns a symmetric NES-versus-adjusted-p-value ggplot |
| `plot_gsea_half_volcano()` | returns a directional ggplot after the explicit NES boundary |
| `plot_gsea_nes_scatter()` | returns a ggplot from the exact shared term-key intersection |

Source documentation defines the complete argument contracts, including the common `term_col`, `nes_col`, `pvalue_col`, `padj_col`, and optional `id_col` arguments.

# Reproduce the repository

Install the rendering dependencies and run the maintained scripts from the repository root:

```r
utils::install.packages(c(
  'base64enc', 'bookdown', 'knitr', 'rmarkdown', 'svglite', 'systemfonts'
))
```

```sh
Rscript R/01_generate_gsea_figures.R
Rscript R/02_render_github_pages.R
Rscript R/03_check_gsea_contracts.R
```

The figure script writes six editable, font-embedded SVGs under
`tutorial/assets/figures/`. The site script renders this source into `docs/`,
and the regression script verifies fixed scientific counts, stable-key
matching, and invalid-input boundaries. Local and continuous-integration checks
are documented in `MAINTENANCE.md`.

# References

<ol class="reference-list">
<li>Subramanian A, Tamayo P, Mootha VK, et al. Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. <em>PNAS</em>. 2005. <a href="https://www.pnas.org/doi/10.1073/pnas.0506580102">https://www.pnas.org/doi/10.1073/pnas.0506580102</a></li>
<li>Ciceri G, Baggiolini A, Cho HS, et al. An epigenetic barrier sets the timing of human neuronal maturation. <em>Nature</em>. 2024;626:881-890. <a href="https://doi.org/10.1038/s41586-023-06984-8">https://doi.org/10.1038/s41586-023-06984-8</a></li>
<li>Xu N, Cho HS, Hackland JOS, et al. Genome-wide CRISPR screen identifies Menin and SUZ12 as regulators of human developmental timing. <em>Nature Cell Biology</em>. 2025;27:1411-1421. <a href="https://doi.org/10.1038/s41556-025-01751-5">https://doi.org/10.1038/s41556-025-01751-5</a></li>
<li>Vuong CK, Weber A, Seong P, et al. A single-cell multiomic analysis identifies molecular and gene-regulatory mechanisms dysregulated in developing Down syndrome neocortex. <em>Science</em>. 2026;392:eaea1259. <a href="https://doi.org/10.1126/science.aea1259">https://doi.org/10.1126/science.aea1259</a></li>
<li>Risgaard RD, et al. Molecular and cellular processes disrupted in the early postnatal Down syndrome prefrontal cortex. <em>Science</em>. 2026;392:eaea1549. <a href="https://doi.org/10.1126/science.aea1549">https://doi.org/10.1126/science.aea1549</a></li>
</ol>
