---
title: "Comparative ranked GSEA waterfall, volcano, scatter-style plots"
description: "Standalone R functions for plotting precomputed GSEA results."
site: bookdown::bookdown_site
documentclass: book
---

```{r tutorial-options, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = 'center'
)
```

<div class="tutorial-title-block">
<h1 class="title">Comparative ranked GSEA waterfall, volcano, scatter-style plots</h1>
<div class="tutorial-subtitle">Standalone R functions for ranked GSEA GO terms, directional volcano plots, and cross-contrast NES comparisons.</div>

<div class="tutorial-meta-block">
<div><span>Author:</span> Zoheb Khan</div>
<div><span>Example:</span> GSE122380 GO biological-process GSEA contrasts</div>
<div><span>Source:</span> <a href="https://github.com/ZohebKhan1/gsea-waterfall">ZohebKhan1/gsea-waterfall</a></div>
</div>
</div>

# Overview

`gsea-waterfall` is a copyable R plotting toolkit for **precomputed** Gene Set Enrichment Analysis (GSEA) results. It provides ranked positive and negative NES waterfall plots, a symmetric volcano plot, directional half-volcano plots, and comparative NES scatterplots.

The functions do not run GSEA or modify supplied enrichment statistics. They select and display rows according to explicit plotting arguments. The worked example uses GO biological-process results from GSE122380 cardiomyocyte differentiation contrasts.

Rendered tutorial: <https://zohebkhan1.github.io/gsea-waterfall/>

The waterfall layout is inspired by pathway-level figures in [Ciceri, *Nature*, 2024](https://doi.org/10.1038/s41586-023-06984-8), [Xu, *Nature Cell Biology*, 2025](https://doi.org/10.1038/s41556-025-01751-5), [Vuong, *Science*, 2026](https://doi.org/10.1126/science.aea1259), and [Risgaard, *Science*, 2026](https://doi.org/10.1126/science.aea1549).

# Install And Source The Functions

Install the plotting dependencies once:

```r
utils::install.packages(c('ggplot2', 'ggrepel', 'RColorBrewer', 'scales'))
```

Download the shared utilities first, then the plot modules you need:

```r
base_url <- 'https://raw.githubusercontent.com/ZohebKhan1/gsea-waterfall/main/functions'

utils::download.file(paste0(base_url, '/gsea_plot_utils.R'), 'gsea_plot_utils.R')
utils::download.file(paste0(base_url, '/gsea_waterfall_plot.R'), 'gsea_waterfall_plot.R')
utils::download.file(paste0(base_url, '/gsea_volcano_plot.R'), 'gsea_volcano_plot.R')
utils::download.file(paste0(base_url, '/gsea_half_volcano_plot.R'), 'gsea_half_volcano_plot.R')
utils::download.file(paste0(base_url, '/gsea_nes_scatter_plot.R'), 'gsea_nes_scatter_plot.R')

source('gsea_plot_utils.R')
source('gsea_waterfall_plot.R')
source('gsea_volcano_plot.R')
source('gsea_half_volcano_plot.R')
source('gsea_nes_scatter_plot.R')
```

# Input Contract

Each table represents one GSEA contrast with one row per unique term. The default columns are:

| Column | Required | Meaning |
|:---|:---|:---|
| `go_description` | yes | unique GO term name and default matching key |
| `NES` | yes | normalized enrichment score |
| `pval` | yes | nominal GSEA p-value in `[0, 1]` |
| `padj` | yes | adjusted p-value in `[0, 1]` |

All four columns must be complete, and NES values must be finite. If descriptions are not unique, supply a complete unique identifier column through `id_col`. Comparative scatterplots use the exact shared identifier intersection; they do not impute absent terms.

Load the bundled examples from the repository root:

```r
gsea_cardiomyocyte_vs_mesoderm <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_cardiomyocyte_vs_mesoderm.csv'
)
gsea_day9_vs_day6 <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_day9_vs_day6.csv'
)
gsea_day3_vs_day1 <- read_gsea_result_csv(
  'tutorial/data/GSE122380_gsea_day3_vs_day1.csv'
)
```

# Waterfall Plots

`plot_gsea_waterfall()` selects one NES direction, ranks terms by `NES`, `padj`, or `pvalue`, and displays the first `top_n` rows. `term_groups` controls point colors; `label_terms` independently controls direct labels. Text seeds are matched case-insensitively against term descriptions, while supplied term IDs are matched exactly after case normalization.

## Positive NES waterfall

```r
positive_term_groups <- list(
  'Ion channel' = c(
    'Calcium', 'Sodium', 'Potassium', 'Ion', 'Voltage', 'Channel',
    'electrical', 'Action potential', 'transmembrane', 'conduction'
  ),
  'Metabolism' = c(
    'Metabolic', 'Metabolism', 'Electron transport chain', 'Mitochondria',
    'Mitochondrial', 'Oxidative', 'Phosphorylation', 'ATP', 'lipid',
    'biosynthetic', 'respiration'
  ),
  'Muscle contraction' = c(
    'Muscle contraction', 'Muscle cell differentiation', 'Striated',
    'Sarcomere', 'Myofibril', 'Actin', 'Muscle'
  ),
  'Heart development' = c('atrial', 'ventricular', 'heart', 'cardiac')
)

positive_label_terms <- c(
  'myofibril assembly',
  'cardiac muscle contraction',
  'cardiac muscle cell action potential',
  'calcium-mediated signaling',
  'electron transport chain',
  'muscle cell differentiation',
  'ATP metabolic process',
  'cardiac cell development',
  'cardiac chamber morphogenesis',
  'muscle contraction',
  'cardiac conduction',
  'cardiac muscle cell contraction',
  'heart growth',
  'cardiac atrium development',
  'heart trabecula morphogenesis'
)

positive_waterfall <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'positive',
  top_n = 100L,
  rank_by = 'NES',
  x_label = 'Ranked GO terms by NES for Cardiomyocyte vs. Mesoderm',
  label_terms = positive_label_terms,
  term_groups = positive_term_groups
)
```

```{r positive-waterfall-figure, echo=FALSE, results='asis', out.width='80%', fig.align='center'}
cat('<div id="fig-positive-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_positive.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 1.</strong> Positive NES GO biological-process waterfall for Cardiomyocyte vs. Mesoderm. Terms are ranked by NES, colored by manually defined cardiac-relevant groups, and labeled with selected terms.</div>

## Negative NES waterfall

```r
negative_term_groups <- list(
  'Ribosomal' = c('ribosome', 'ribosomal', 'spliceosomal', 'rRNA', 'RNA'),
  'Mitosis' = c(
    'Mitosis', 'Meiosis', 'Nuclear division', 'cell cycle', 'telomere',
    'spindle', 'Mitotic', 'Meiotic'
  ),
  'DNA replication' = c('chromosome', 'DNA', 'base', 'repair', 'DNA-')
)

negative_term_group_colors <- c(
  'Ribosomal' = '#377EB8',
  'Mitosis' = '#984EA3',
  'DNA replication' = '#E41A1C'
)

negative_waterfall <- plot_gsea_waterfall(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'negative',
  top_n = 100L,
  rank_by = 'NES',
  x_label = 'Ranked GO terms by NES for Cardiomyocyte vs. Mesoderm',
  label_n = 10L,
  term_groups = negative_term_groups,
  term_group_colors = negative_term_group_colors
)
```

```{r negative-waterfall-figure, echo=FALSE, results='asis', out.width='80%', fig.align='center'}
cat('<div id="fig-negative-waterfall" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_waterfall_cardiomyocyte_vs_mesoderm_negative.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 2.</strong> Negative NES GO biological-process waterfall for the same contrast. The selected groups highlight ribosomal, mitotic, and DNA-replication programs.</div>

# Volcano Plots

Use `plot_gsea_volcano()` for both NES directions in one panel and `plot_gsea_half_volcano()` for a larger directional view. The example retains the established adjusted p-value cutoff of 0.05.

## Symmetric volcano

```r
volcano_label_terms <- c(
  'cardiac muscle contraction',
  'muscle cell differentiation',
  'cardiac chamber morphogenesis',
  'heart growth',
  'heart valve development',
  'mitotic nuclear division',
  'DNA replication initiation',
  'mitotic G2/M transition checkpoint',
  'centriole assembly',
  'ribosome assembly'
)

symmetric_volcano <- plot_gsea_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  padj_cutoff = 0.05,
  label_terms = volcano_label_terms,
  contrast_label = 'Cardiomyocyte vs. Mesoderm',
  label_words_per_line = 3L
)
```

```{r symmetric-volcano-figure, echo=FALSE, results='asis', out.width='58%', fig.align='center'}
cat('<div id="fig-symmetric-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_volcano_cardiomyocyte_vs_mesoderm.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 3.</strong> Symmetric GO biological-process GSEA volcano for Cardiomyocyte vs. Mesoderm. Point color marks significant positive, significant negative, or non-significant enrichment.</div>

## Directional half-volcano plots

```r
positive_half_volcano_label_terms <- c(
  'myofibril assembly',
  'cardiac muscle contraction',
  'cardiac muscle cell action potential',
  'electron transport chain',
  'muscle cell differentiation',
  'cardiac cell development',
  'cardiac chamber morphogenesis',
  'muscle contraction',
  'cardiac conduction',
  'heart growth',
  'heart trabecula morphogenesis'
)

positive_half_volcano <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'positive',
  p_col = 'padj',
  padj_cutoff = 0.05,
  label_terms = positive_half_volcano_label_terms,
  term_groups = positive_term_groups,
  inner_nes_limit = 1,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

```{r positive-half-volcano-figure, echo=FALSE, results='asis', out.width='72%', fig.align='center'}
cat('<div id="fig-positive-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_positive.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 4.</strong> Positive NES half-volcano for Cardiomyocyte vs. Mesoderm. The enlarged view uses the same biological grouping logic as the waterfall plot.</div>

```r
negative_half_volcano_label_terms <- c(
  'DNA-templated DNA replication',
  'DNA replication',
  'cell cycle checkpoint signaling',
  'mitotic cell cycle checkpoint signaling',
  'ribosome biogenesis',
  'recombinational repair',
  'DNA replication initiation',
  'cell cycle DNA replication',
  'spindle organization'
)

negative_half_volcano <- plot_gsea_half_volcano(
  gsea_results = gsea_cardiomyocyte_vs_mesoderm,
  direction = 'negative',
  p_col = 'padj',
  padj_cutoff = 0.05,
  label_terms = negative_half_volcano_label_terms,
  term_groups = negative_term_groups,
  term_group_colors = negative_term_group_colors,
  inner_nes_limit = 1,
  contrast_label = 'Cardiomyocyte vs. Mesoderm'
)
```

```{r negative-half-volcano-figure, echo=FALSE, results='asis', out.width='72%', fig.align='center'}
cat('<div id="fig-negative-half-volcano" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_half_volcano_cardiomyocyte_vs_mesoderm_negative.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 5.</strong> Negative NES half-volcano for Cardiomyocyte vs. Mesoderm. Selected labels emphasize DNA replication, checkpoint, and ribosome-associated programs.</div>

# Comparative NES Scatterplot

`plot_gsea_nes_scatter()` compares the exact shared term-key intersection. The fitted line is descriptive; the helper does not calculate a new enrichment statistic.

```r
scatter_label_terms <- c(
  'mesenchyme development',
  'mesoderm development',
  'mesodermal cell differentiation',
  'myofibril assembly',
  'heart process',
  'oxidative phosphorylation',
  'regulation of nuclear division'
)

all_quadrants_scatter <- plot_gsea_nes_scatter(
  gsea_x = gsea_day9_vs_day6,
  gsea_y = gsea_day3_vs_day1,
  x_label = 'Day 9 vs. Day 6 NES',
  y_label = 'Day 3 vs. Day 1 NES',
  color_by = 'significance',
  quadrant = 'all',
  x_name = 'Day 9 vs. Day 6',
  y_name = 'Day 3 vs. Day 1',
  label_terms = scatter_label_terms,
  padj_cutoff = 0.05,
  include_nonsignificant = TRUE,
  equal_axis_limits = TRUE,
  show_fit_line = TRUE
)
```

```{r all-quadrants-scatter-figure, echo=FALSE, results='asis', out.width='100%', fig.align='center'}
cat('<div id="fig-all-quadrants-scatter" class="figure-anchor"></div>\n')
knitr::include_graphics('assets/figures/GSE122380_gsea_scatter_day9_vs_day6_x_day3_vs_day1_all_quadrants.svg')
```

<div class="tutorial-figure-caption"><strong>Figure 6.</strong> Comparative GO biological-process NES scatterplot for Day 9 vs. Day 6 and Day 3 vs. Day 1. Point color shows significance in one, both, or neither contrast.</div>

# Function Reference

| Function | Contract |
|:---|:---|
| `read_gsea_result_csv()` | reads and validates one precomputed CSV without filtering rows |
| `plot_gsea_waterfall()` | returns a ranked one-direction ggplot |
| `plot_gsea_volcano()` | returns a symmetric NES-versus-adjusted-p-value ggplot |
| `plot_gsea_half_volcano()` | returns a directional ggplot after the explicit NES boundary |
| `plot_gsea_nes_scatter()` | returns a ggplot from the exact shared term-key intersection |

All plot functions return ggplot objects and do not write files. Their source-level documentation defines the complete argument and return contracts. The common input-column arguments are `term_col`, `nes_col`, `pvalue_col`, `padj_col`, and optional `id_col`; visual arguments control labels, colors, dimensions inside the panel, and legends without changing the supplied enrichment values.

# Reproduce The Repository

Install the rendering dependencies and run the maintained scripts from the repository root:

```r
utils::install.packages(c(
  'base64enc', 'bookdown', 'knitr', 'rmarkdown', 'svglite', 'systemfonts'
))
```

```sh
Rscript R/01_generate_gsea_figs.R
Rscript R/02_render_github_pgs.R
Rscript R/03_check_gsea_contracts.R
```

The first script writes the six editable, font-embedded SVGs under `tutorial/assets/figures/`. The second renders `tutorial/tutorial.Rmd` into `docs/` for GitHub Pages. The third verifies the fixed scientific counts, plot contracts, public API, identifier matching, and invalid-input boundaries. Maintainers use the canonical R Codex Utils commands recorded in `MAINTENANCE.md`.

# References

<ol class="reference-list">
<li>Subramanian A, Tamayo P, Mootha VK, et al. Gene set enrichment analysis: a knowledge-based approach for interpreting genome-wide expression profiles. <em>PNAS</em>. 2005. <a href="https://www.pnas.org/doi/10.1073/pnas.0506580102">https://www.pnas.org/doi/10.1073/pnas.0506580102</a></li>
<li>Korotkevich G, Sukhov V, Budin N, Shpak B, Artyomov MN, Sergushichev A. Fast gene set enrichment analysis. <a href="https://www.biorxiv.org/content/10.1101/060012v3">https://www.biorxiv.org/content/10.1101/060012v3</a></li>
<li>Bioconductor `fgsea`: Fast Gene Set Enrichment Analysis. <a href="https://bioconductor.org/packages/fgsea/">https://bioconductor.org/packages/fgsea/</a></li>
<li>The Gene Ontology Consortium. The Gene Ontology resource. <a href="https://geneontology.org/">https://geneontology.org/</a></li>
<li>Ciceri G, Baggiolini A, Cho HS, et al. An epigenetic barrier sets the timing of human neuronal maturation. <em>Nature</em>. 2024;626:881-890. <a href="https://doi.org/10.1038/s41586-023-06984-8">https://doi.org/10.1038/s41586-023-06984-8</a></li>
<li>Xu N, Cho HS, Hackland JOS, et al. Genome-wide CRISPR screen identifies Menin and SUZ12 as regulators of human developmental timing. <em>Nature Cell Biology</em>. 2025;27:1411-1421. <a href="https://doi.org/10.1038/s41556-025-01751-5">https://doi.org/10.1038/s41556-025-01751-5</a></li>
<li>Vuong CK, Weber A, Seong P, et al. A single-cell multiomic analysis identifies molecular and gene-regulatory mechanisms dysregulated in developing Down syndrome neocortex. <em>Science</em>. 2026;392:eaea1259. <a href="https://doi.org/10.1126/science.aea1259">https://doi.org/10.1126/science.aea1259</a></li>
<li>Risgaard RD, et al. Molecular and cellular processes disrupted in the early postnatal Down syndrome prefrontal cortex. <em>Science</em>. 2026;392:eaea1549. <a href="https://doi.org/10.1126/science.aea1549">https://doi.org/10.1126/science.aea1549</a></li>
</ol>
